Infant Birth Weight in Brazil

### 2024-08-15 products/manuscript/Manuscript_Final_Lancet.qmd
Revisions for submission 2
After call w/Bill

## Methods
This cross-sectional analysis uses hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital. I used linear models to assess differences in infant birth weight by maternal skin color, age, parity, and nationality.

## Findings
Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value <0.01) than infants born to Euro-descendant women, whether Brazilian or immigrant. Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 96 grams lighter (p-value <0.01) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value <0.05) than infants born to White women.

@@

# Introduction
Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004]. In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research, as LBW is associated with higher rates of neonatal and infant mortality and possible health problems over the life course [@cardoso2005; @paixao2021; @vilanova2019].

Contemporary studies in Brazil have shown that non-White mothers have higher rates of LBW [@nyarko2013; @wehby2015a; @vilanova2019]. These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns have higher rates of LBW than White newborns, with trends worsening over time [@pollock2021]. Racism is a major factor in these outcomes, as studies demonstrate that U.S.-born Black women have higher rates of LBW and PTB than Black immigrant women [@braveman2021a; @clay2023a; @david1997].

@@

Yet recent studies also show an overall improvement in human welfare during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019]. The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe. Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@benchimol1999; @hochman2013]. Although infant mortality rates declined from the nineteenth century, the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@roth2020]. Worldwide, the combination of these advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992; @woods2009].

@@

@@

In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon. In our current postgenomic era, in which race as a biological category has been rejected, the possibility for the reaffirmation of centuries-old patterns of hierarchical scientific thinking remains very real [@roberts2011]. Contemporary research on epigenetic markers stemming from common experiences can create essentialist racial thinking just as past, debunked research into any genetic existence of race did [@meloni2022]. The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not "pathologically" different or part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].

This study analyzes infant birth weight as a proxy for general fetal and neonatal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras. It aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.

@@

This analysis draws on a unique sample of 2,845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926. I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO). The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). In the initial months of publication, the reports included more complete information, including labor and delivery time or detailed descriptions of surgical procedures. Over time, clinical notes became streamlined. When physicians intervened in birth, the notes included the type of intervention, the indication, and the attending obstetrician [@roth2020; @roth2020a].

@@

*Birth weight*. The outcome was birth weight measured as a continuous variable in grams for live births only. In a secondary analysis, I also evaluated birth weight as a binary variable (either normal birth weight, NBW (<p>&ge;</p>2500 g) or LBW (<2500 g), according to current WHO classifications) [@worldhealthorganization2022].

@@

Racial terminology in Brazil was, and continues to be, complex and dynamic [@telles2006; @travassos2004]. Racial categories and skin color exist on a spectrum rather than a Black-White binary [@telles2006]. Analyzing both specific categories of skin color and more general categories of ancestry can help address these different understandings of race and is also in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].

@@

*Maternal outcome*. Hospital records recorded maternal outcome was recorded as death, discharge, or hospital transferal, which I maintain here. I assume that hospital transferal did not result in maternal death in my calculations of maternal mortality below.

@@

I used Ordinary Least Squares (OLS) to estimate the relationship of maternal variables on infant birth weight. I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019]. Due to the lack of comprehensive information on gestational age in these data, I did not consider birth weight in reference to gestational age, as is best practice for disentangling PTB from LBW [@cutland2017]. I discuss this constraint in the discussion below.

I further ran logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable, in line with some studies on birth weight [@falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a]. Because most results were not statistically significant in these models, I do not include adjusted models here. Unadjusted and adjusted logistic models and results are included in the supplemental analysis.

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term.

The simple linear regression model for original color is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term. Results are displayed in Table 4.

@@

$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term.

@@

The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers. Results for all linear models are displayed in Tables 2 and 3. All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race). When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 85 grams lighter than their Euro-descended counterparts (Table 2). In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 100 grams less than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter than White mothers. Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of only European descent.

@@

Unadjusted ORs (Table 4) demonstrate that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.43, 95% CI 1.07, 1.95, p-value 0.019). When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were not significant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3).

@@

Discussions of epigenetic markers associated with the lived experience of racism might also be applicable here. Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors), mediated through epigenetic mechanisms. Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance [@leimert2020]. Yet, it is crucial to remember that these are not deterministic processes but can be counteracted by social and political interventions that address the root causes of racism and its effects on health [@meloni2022].

@@

*Conclusion*
Lack of data on maternal nutrition and health status, which cannot be corrected for by using maternal skin color as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study. Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.

@@

```{r}
#| label: tbl-table2
#| tbl-cap: "Model 1 and 2 Results"
#| echo: FALSE
table2 <- readRDS(here::here("results", "tables", "table3_final.rds"))
flex_table2 <- as_flex_table(table2)
flex_table2 <- flex_table2 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table2
```
{{< pagebreak >}}

```{r}
#| label: tbl-table3
#| tbl-cap: "Model 3 and 4 Results"
#| echo: FALSE
table3 <- readRDS(here::here("results", "tables", "table4_final.rds"))
flex_table3 <- as_flex_table(table3)
flex_table3 <- flex_table3 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table3
```
{{< pagebreak >}}

>>>>

## Methods
This cross-sectional analysis uses hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital. I used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.

## Findings
African ancestry was correlated with lower birth weights. Infants born to Afro-descendant women had birth weights estimated to be 85 grams lighter (p-value <0.01) than infants born to Euro-descendant women, whether Brazilian or immigrant. Among Afro-descendant women, infants born to Black (*preta*) women had birth weights estimated to be 100 grams lighter (p-value <0.01) than infants born to White women, and infants born to mixed-race (*parda*) women had birth weights estimated to be 72 grams lighter (p-value <0.05) than infants born to White women.

@@

# Introduction
Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004]. In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research, as LBW is associated with higher rates of neonatal and infant mortality and possible health problems over the life course [@cardoso2005; @paixao2021; @vilanova2019]. BILL COMMENT: ARE THERE LONGITUDINAL STUDIES OF LATER-LIFE MORTALITY OR LIFE EXPECTANCY WITH BIRTH WEIGHT?

Contemporary studies in Brazil have shown that non-White mothers have higher rates of LBW [@nyarko2013; @wehby2015a; @vilanova2019]. These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns have higher rates of LBW than White newborns, with trends worsening over time [@pollock2021]. Racism is a major factor in these outcomes, as studies demonstrate that U.S.-born Black women have higher rates of LBW and PTB than Black immigrant women [@braveman2021a; @clay2023a; @david1997]. BILL COMMENT: RELATED TO RACISM, WHAT DOES LITERATURE ON BIRTH WEIGHT SAY ABOUT ITS CAUSES? THESE ARE CAUSAL STATEMENTS.

@@

Yet recent studies also show an overall improvement in human welfare for those who survived infancy during this period, measured in increased height at the population level, particularly in Brazil's southeastern region, where Rio de Janeiro is located [@franken2019]. The early twentieth century was a period of advancement in both the provision of clinical care and the implementation of public health initiatives in Brazil and across the globe. Improved sanitation measures helped stem infectious disease outbreaks and improve quality of life [@benchimol1999; @hochman2013]. Although infant mortality rates declined from the nineteenth century, the major advancements in clinical medicine that provided a sustained improvement in maternal mortality, stillbirth, and infant mortality rates, including blood transfusions and antibiotics, did not come about until the late 1930s and early 1940s [@roth2020]. Worldwide, the combination of these advancements in medical care and overall improvements in nutrition resulted in dramatic and sustained drops in maternal and infant mortality after World War II [@loudon1992; @woods2009].

@@

BILL COMMENT: I AM CURIOUS, DO WHITE SOUTH AFRICANS HAVE HIGHER BIRTH WEIGHT, CONTROLLING FOR INCOME? ANY STUDY OF THIS?

@@

In exploring differences in infant birth weight according to maternal skin color in early twentieth-century Brazil, my aim is not to reify race as a genetic or epigenetic phenomenon. In our current postgenomic era, in which race as a biological category has been rejected, the possibility for the reaffirmation of centuries-old patterns of hierarchical scientific thinking remains very real [@meloni2022]. Contemporary research on epigenetic markers stemming from common experiences can create essentialist racial thinking just as past, debunked research into any genetic existence of race did [@meloni2022]. The working-class women and their children who relied on Laranjeiras for prenatal and delivery care were not "pathologically" different or part of permanently "damaged" communities (socially or biologically) but rather were living the consequences of centuries of chattel slavery and its aftermath [@tuck2009].

This study analyzes infant birth weight as a proxy for general fetal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras. It aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives. BILL COMMENT: PUT THIS PARAGRAPH MUCH EARLIER.

@@

This analysis draws on a unique data set of 2,845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926. I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO). The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). In the initial months of publication, the reports included more complete information, including labor and delivery time or detailed descriptions of surgical procedures. Over time, clinical notes became streamlined. When physicians intervened in birth, the notes included the type of intervention, the indication, and the attending obstetrician [@roth2020; @roth2020a]. BILL COMMENT: FOR OBSTETRICIANS WHO APPEAR MORE THAN ONCE, IS THERE A RELATIONSHIP WITH BIRTH WEIGHT?

@@

*Birth weight*. The outcome was birth weight measured as a continuous variable in grams for live births only. In a secondary analysis, I also evaluated birth weight as a binary variable (either normal birth weight, NBW (<p>&ge;</p>2500 g) or LBW (<2500 g), according to current WHO classifications) [@worldhealthorganization2022]. BILL COMMENT: THESE COME FROM WHERE AND WHEN?

@@

Except at the extremes, racial classification in Brazil was, and continues to be, imprecise [@travassos2004]. Racial categories and skin color exist on a spectrum rather than a Black-White binary [@telles2006]. Using more general categories addresses imprecision by grouping all possible mothers of any African descent into one category. This method is in line with recent studies, which look at all African-descended peoples as a group and then stratify by racial mixing [@nyarko2013; @wehby2015a].

@@

*Maternal outcome*. Hospital records recorded maternal outcome as death, discharge, or hospital transferal, which I maintain here. I assume that hospital transferal did not result in maternal death in my calculations of maternal mortality below.

@@

I used Ordinary Least Squares (OLS) to establish a relationship of maternal variables and infant birth weight. I employ birth weight as a continuous variable in line with both current and historical studies [@andrasfay2020a; @costa1998; @costa2004; @david1997; @leal2006; @silveira2019]. Due to the lack of comprehensive information on gestational age in these data, I did not consider birth weight in reference to gestational age, as is best practice for disentangling PTB from LBW [@cutland2017]. I discuss this constraint in the discussion below.

I further ran logistic models that estimated the relationship of maternal variables on infant birth weight as a categorical variable, in line with some studies on birth weight [@falcao2020; @nyarko2013; @paixao2021; @silva1998; @wehby2015a; @wehby2017a]. Because most results were not statistically significant in these models, I do not include adjusted models here. Unadjusted and adjusted logistic models and results are included in the supplemental analysis. BILL COMMENT: WHY WEREN'T MODELS SIGNIFICANT?

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), and $\epsilon$ is the error term.

@@

$$\text{BW} = \alpha + \beta_1 \text{MA} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MA` is maternal ancestry (Euro-descent, Afro-descent), `Gest` is gestational status (nulliparous or multiparous), `Age` is maternal age in years, and $\epsilon$ is the error term. Results for both models are displayed in Table 2.

The simple linear regression model for original skin color is as follows:

$$\text{BW} = \alpha + \beta_1 \text{MC} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_1$ is the group deviation from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), and $\epsilon$ is the error term.

@@

$$\text{BW} = \alpha + \beta_1 \text{MC} + \beta_2 \text{Gest} + \beta_3 \text{Nat} + \beta_4 \text{Age} + \epsilon$$ whereas $\alpha$ is the average birth weight, $\beta_x$ are the group deviations from the average birth weight, `BW` is birth weight in grams, `MC` is maternal skin color (White, Black, and mixed race), `Gest` is gestational status (nulliparous or multiparous), `Nat` is maternal nationality (Brazilian, Latin American, European, Middle Eastern), `Age` is maternal age in years, and $\epsilon$ is the error term. Results for both models are displayed in Table 3.

@@

The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers. A difference in means test shows borderline significance between the means (t-statistic = 1.9119, df. 2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]).

Results for all linear models are displayed in Tables 2 and 3. All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race). When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 85 grams lighter than their Euro-descended counterparts (Table 2). In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 100 grams less than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter than White mothers. Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of only European descent. AT THIS POINT I NEED TO SEE A LITERATURE THAT TELLS ME 100 GRAMS MATTERS FOR SOMETHING, HEALTH-WISE, IS 100 GRAMS MORE THAN 2 STANDARD DEVIATIONS?

@@

Unadjusted ORs (Table 4) demonstrate that women of African-descent had higher odds of giving birth to infants with lower birth weight than Euro-descended women (OR 1.43, 95% CI 1.07, 1.95, p-value 0.019). When using the hospital records' original skin color categories, mixed-race women had higher odds of giving birth to infants of lower birth weight (OR 1.64, 95% CI	1.17, 2.30, p-value 0.004), but results for the infants of Black women were not significant (OR 1.23, 95% CI 0.85, 1.77, p-value 0.3). MAKING CONTINUOUS VARIBLE CATEGORICAL. LBW NOT A CATEGORY YET PER WHO.

@@

Discussions of epigenetic markers associated with the lived experience of racism might also be applicable here. Studies have connected racial disparities in pregnancy outcomes, including PTB and LBW, with the accumulation of stress over the pregnant person’s life course (and the life courses of their ancestors) understood as allostatic load measured by a variety of biomarkers such as cortisol and blood pressure, mediated through epigenetic mechanisms. Although there is no genetic basis for race, experiences of racism can result in the accumulation of stressors that can be expressed epigenetically and lead to adverse pregnancy and health outcomes across generations through transgenerational epigenetic inheritance [@leimert2020]. Yet, it is crucial to remember that these are not deterministic processes but can be counteracted by social and political interventions that address the root causes of racism and its effects on health [@meloni2022].

@@

Lack of data on maternal nutrition and health status, which cannot be corrected for by using maternal skin color as a proxy for overall health, is a major impediment to understanding the true relationship between maternal variables and infant birth weight in this study. Given the paucity of historical understandings of current-day health disparities in maternal-infant health, this paper demonstrates the need for more research into quantitative associations between socio-demographic categories and health in the past.

*Conclusion*
This study has shown that birth weight differed by maternal skin color. This suggests that racial inequality stemming from slavery impacted health conditions in the post-abolition period, possibly through multiple mechanisms. These findings, although not generalizable beyond this specific hospital, are consistent with present-day studies, demonstrating how racial inequality in health outcomes has persisted over time.

@@

```{r}
#| label: tbl-table2
#| tbl-cap: "Model 1 and 2 Results (Dependent variable: Birth weight in grams)"
#| echo: FALSE
table2 <- readRDS(here::here("results", "tables", "table3_final.rds"))
flex_table2 <- as_flex_table(table2)
flex_table2 <- flex_table2 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table2
```
{{< pagebreak >}}

```{r}
#| label: tbl-table3
#| tbl-cap: "Model 3 and 4 Results (Dependent variable: Birth weight in grams)"
#| echo: FALSE
table3 <- readRDS(here::here("results", "tables", "table4_final.rds"))
flex_table3 <- as_flex_table(table3)
flex_table3 <- flex_table3 %>%
  flextable::set_table_properties(width = .5, align = "center")
flex_table3
```
{{< pagebreak >}}
